Geographically aware estimates of remotely sensed water properties for Chesapeake Bay

### 2022-11-17 manuscript/manuscript.tex
text edits

Unfortunately, direct observations alone are not adequate for many applications that require spatially extensive data such as \hl{Sea Level Rise (SLR)} forecasting and habitat suitability assessments \cite{hoodChesapeakeBayProgram2021}. For example, habitat suitability assessments are typically done on a gridded basis for the entirety of a particular estuary rather than at selected point locations. Applications such as these often turn to gridded alternatives to direct observations such as output from directly-observing satellite-based sensors \cite{foreCombinedActivePassive2016} or operational now/fore-casting products \cite{lanerolle2011second}. Unfortunately, these too are of limited use for applications such as validation of next-generation Earth Systems Models (ESMs, \cite{golazDOEE3SMModel2022}), where the temporal availability of operational forecasting products is too short ($<$ 1 year) and the spatial resolution of directly-observing satellite-based sensors is too coarse ($>$ 30 $km^2$).

@@

We supplemented observational data with a grid of “freshwater influence” (FWI) estimates where each grid cell represents the average of annual-flow weighted distance of that cell from the 5 major tributaries (Susquehanna, James, Pautexent, Choptank, and Potomac). Annual flow magnitudes were calculated from National Water Information System instantaneous data (sites: 01488110, 01491000, 01576000, 01654000, 01594440, 02037500, \cite{nwis2021usgs}). Rather than straight-line distances we used “in-water” distances calculated using the scikit-image Python package \cite{littleKrigingEstuariesCrow1997,vanderwaltScikitimageImageProcessing2014}. Our in-water distance calculations involved creating a cost–surface whereby land pixels had a very high cost relative to water pixels. Then the distance from any one cell to a tributary source was the accumulated travel “cost” as given by the scikit-image \texttt{route\_through\_array} function. We combined FWI estimates with the pixel-wise daily mean observations and regressed them in RF models against data from the \hl{1000m} resolution MODIS Aqua product, which provides \hl{9} bands of values for each pixel spanning from \hl{the following wavelength ranges (405-420, 438-448, 483-493, 526-536, 546-556, 662-672, 673-683, 743-753, and 862-877nm)} \cite{vermoteericMOD09GAMODISTerra2015}. \hl{MODIS data was pulled from Google Earth Engine (ID:MODIS/006/MYDOCGA) using the \texttt{earthengine-api} Python package.}

@@

>>>>

Unfortunately, direct observations alone are not adequate for many applications that require spatially extensive data such as \hl{Sea Level Rise (SLR)} forecasting and habitat suitability assessments \cite{hoodChesapeakeBayProgram2021}. For example, habitat suitability assessments are typically done on a gridded basis for the entirety of a particular estuary rather than at selected point locations. Applications such as these often turn to gridded alternatives to direct observations such as output from directly-observing satellite-based sensors \cite{foreCombinedActivePassive2016} or operational now/fore-casting products \cite{lanerolle2011second}. Unfortunately, these too are of limited use for applications such as validation of next-generation Earth Systems Models (ESMs, \cite{golazDOEE3SMModel2022, xuDevelopmentLandRiver2022a}), where the temporal availability of operational forecasting products is too short ($<$ 1 year) and the spatial resolution of directly-observing satellite-based sensors is too coarse ($>$ 30 $km^2$).

@@

We supplemented observational data with a grid of “freshwater influence” (FWI) estimates where each grid cell represents the average of annual-flow weighted distance of that cell from the 5 major tributaries (Susquehanna, James, Pautexent, Choptank, and Potomac). Annual flow magnitudes were calculated from National Water Information System instantaneous data (sites: 01488110, 01491000, 01576000, 01654000, 01594440, 02037500, \cite{nwis2021usgs}). Rather than straight-line distances we used “in-water” distances calculated using the scikit-image Python package \cite{littleKrigingEstuariesCrow1997,vanderwaltScikitimageImageProcessing2014}. Our in-water distance calculations involved creating a cost–surface whereby land pixels had a very high cost relative to water pixels. Then the distance from any one cell to a tributary source was the accumulated travel “cost” as given by the scikit-image \texttt{route\_through\_array} function. We combined FWI estimates with the pixel-wise daily mean observations and regressed them in RF models against data from the \hl{1000m} resolution MODIS Aqua product, which provides \hl{9} bands of values for each pixel spanning from \hl{the following wavelength ranges (405-420, 438-448, 483-493, 526-536, 546-556, 662-672, 673-683, 743-753, and 862-877nm)} \cite{vermoteericMOD09GAMODISTerra2015}. \hl{MODIS data was pulled from Google Earth Engine (Asset: \texttt{MODIS/006/MYDOCGA}) using the \texttt{earthengine-api} Python package.}

@@

\hl{Across all models, MODIS band 8 was the most important reflectance feature. This result may be partly explained by the fact that Band 8 is the shortest wavelength (405-420nm) included in our study and is likely related to the concentration of Colored Dissolved Organic Matter (CDOM) the absorption of which declines exponentially with increasing wavelength under simulated and laboratory conditions} \cite{loiselleOpticalCharacterizationChromophoric2009, sathyendranath2000remote}). \hl{CDOM itself is strongly related to salinity} \cite{sathyendranath2000remote, vogelAssessingSatelliteSea2016, wangDevelopmentMODISData2018} \hl{and thus may be related to other properties such as temperature and turbidity which also vary spatially with the upstream-downstream environmental gradient of estuaries. In field conditions and in remote sensing applications, the absoption spectra of CDOM and the shape of the CDOM-salinity relationship is highly variable from region-to-region as a function of local factors such as bottom reflectance and phytoplankton community composition} \cite{sathyendranath2000remote}. \hl{Although band 8 was not universally the most important feature in prior remote sensing studies of estuarine water properties, the general importance of short wavelength bands is a result shared by prior studies on estuarine salinity} \cite{wangDevelopmentMODISData2018}.
